Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2019-05-23 paper/introduction.tex
jr

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, it can be challenging to identify whether the frequency of a word presented on a subject's retina modulates brain activity because word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Generating a set of words that control for all of these factors simultaneously can be challenging; it is thus common to use forward modeling, i.e. to train a model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from large and complex observations. For example, brain activity is generally recorded with hundreds of sensors simultaneously. It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but they are notoriously difficult to interpret \citep{lebart1995statistique}.

>>>>

First, causal and non-causal factors may be numerous and collinear. In physics, for example, one may be challenged to identify whether fusion is caused by a change in temperature or a change in pressure, as these two factors may, at first, be difficult to manipulate independently. This issue becomes increasingly pronounced as the number of potential factors increases. In neuroscience, for example, identifying whether the frequency of a word presented on a subject's retina modulates brain activity can be surprisingly difficult. Indeed, word frequency covaries with other factors such as their length (short words are more frequent than long words), their categories (determinants are more frequent than adverbs), their individual letters (e.g. words that contain `z' are rare) and so forth \citep{kutas2011thirty,pegado2014timing}. Instead of selecting a set of words that control for all of these factors simultaneously, it is thus common to use forward modeling, i.e. to train a mathematical model to predict observations (e.g. brain activity) from a minimal combination of competing factors (e.g. word length, word frequency, e.g. \citep{huth2016natural}), and investigate, in the model, the estimated contribution of each factor \citep{friston1994statistical}.

The second challenge to measuring causal influence is that observations can be large and complex. The relationship between causes and effects is thus often considered in a backward manner, by training models to maximally predict causes from multidimensional observations. For example, brain activity is often recorded with hundreds or thousands of sensors simultaneously. It is thus common to use `decoding' techniques, by, for example, fitting a support vector machine across multiple sensors to predict the category of a stimulus \citep{cichy2014resolving, king2016brain, kriegeskorte2008representational, norman2006beyond}.

Both forward and backward modeling have competing benefits and drawbacks. Specifically, forward modeling disentangles the independent contribution of collinear factors, but does not combine multidimensional observations. By contrast, backward modeling combines multiple observations, but does not disentangle collinear factors \cite{weichwald2015causal, hebart2018deconstructing, king2018encoding}. There also exist cross-decomposition techniques such as Partial Least Squares (PLS) and Canonical Correlation Analysis (CCA), but their resulting coefficients are notoriously difficult to interpret \citep{lebart1995statistique}.

### 2019-05-23 paper/meg_exps.tex
jr

The raw data was then segmented 100ms before word onset and 1.000 sec after
word onset. Time $t=0$ms corresponds to the word onset. Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240Hz. Figure 1 in the supplementary material
shows the average response of the magnetometers evoked by the words presented
to the first subject.

@@

We attempt to identify which word features ($X$) are reliably associated with the MEG recordings $Y$ at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We applied forward regression (a.k.a encoding coefficients), backward regression (a.k.a decoding scores) and B2B regression on 301-sensor MEG recordings of brain activity, while subjects were flashed with individual words every $\approx$650ms. Top row: the first subject. Bottom row: average across subjects. Each line represent the association between the MEG signals and two word features as a function of time relative to word onset. Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional sum-square operation, which can dimish sensitivity. Backward regression operates across all channels, but does not learn to disambiguate the two word features. B2B efficiently disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
  showing early responses ($\approx$ 150ms) to word length and later responses
  ($\approx$ 400ms) to word frequency.}
  \label{fig:meg_twocurves}
\end{figure}

@@

Figure \ref{fig:meg_twocurves} shows, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after word onset. However,
forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
$t$-value$=18.7$, $p=10^{-31}$, frequency: $t$-value$=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t$-value$=33.6$,
$p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$). However, as backward models confound collinear factors, such high sensitivity likely comes from a loss of specificity. This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present similar dynamics.

>>>>

The raw data was then segmented 100ms before word onset and 1s after
word onset ($t=0$ms corresponds to word onset). Finally, each resulting
segment was baseline-corrected between -100ms and 0ms, and decimated by 5 and
thus led a sampling frequency of 240Hz. Figure 1 in the supplementary material
shows the average response of the magnetometers evoked by the words presented
to the first subject.

@@

Our goal is to identify which word features ($X$) lead to a reliable modulation of brain activity as measured with MEG recordings ($Y$) at each time sample and for each subject separately. To this aim,
we compare three methods: forward regression, backward regression, and back-to-back regression.
Corresponding results can be found in Figure \ref{fig:meg_twocurves}.

\begin{figure}[t!]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/meg_twocurves.pdf}
  \caption{We applied forward regression (a.k.a encoding coefficients), backward regression (a.k.a decoding scores) and B2B regression on 301-sensor MEG recordings of brain activity, while subjects were flashed with individual words every $\approx$650ms. Top row: the first subject. Bottom row: average across subjects. Each line represent the link between word features $X$ and MEG signals $Y$
  as a function of time relative to word onset ($x$ axis). The $y$-axis varies for each model (forward: H coefficients; backward: pearson correlation coefficients; B2B: $\hat E$). Forward regression makes prediction for each MEG channel independently; detecting an effect thus requires an additional average of the norm and can diminish sensitivity. Backward regression operates across all channels, but does not learn to disambiguate the two word features. B2B disambiguates these two factors consistently with previous neuroscientific work \cite{kutas2011thirty}\citep{pegado2014timing}
  showing early responses ($\approx$ 150ms) to word length and later responses
  ($\approx$ 400ms) to word frequency.}
  \label{fig:meg_twocurves}
\end{figure}

@@

Figure \ref{fig:meg_twocurves} shows, for each model, the effects obtained
in a representative subject, as well as the average effects across subjects.
The results show that all models significantly detect that both word length and
word frequency modulate brain responses after word onset. However,
forward regression detected word length ($t$-value$=10.6$, $p=10^{-16}$) and
word frequency ($t$-value$=11.1$, $p=10^-17$) less reliably than B2B (length:
$t$-value$=18.7$, $p=10^{-31}$, frequency: $t$-value$=15.9$, $p=10^{-26}$). B2B
was less sensitive than the backward regression (length: $t$-value$=33.6$,
$p=10^{-50}$, frequency: $t$-value$=31.7$, $p=10^{-48}$). However, as backward models confound collinear factors, this high sensitivity likely comes from a loss of specificity. This possibility is corroborated by the fact that the effects of word length and word frequency identified with backward regression present largely similar dynamics.

### 2019-05-23 paper/supplementary.tex
Typos
typos
jr

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm, clip=True]{figures/ridgecv_baseline_result.pdf}
  \caption{Regression baseline. Coefficients appear to be
  linearly correlated, unlike our method in the main text.}
  \label{fig:ridgebaselineresult}
\end{figure}

>>>>

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues.pdf}
  \caption{We apply a statistical t-test to measure if coefficients were significantly different from baseline value (at word onset, $t=0$ms).
  We compute the negative logarithm of the p-value (\textit{higher is better}: nonzero effects are more easily detected). We plot the average across features, over time. Red: word length, blue: word frequency.
  Full line: B2B, dashed line: Forward regression. We see that our method more easily detects the influence of causes of MEG response.}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth, trim=0cm 0cm 0cm 0cm]{figures/pvalues_vertical.pdf}
  \caption{We show the average value and the standard deviation across subjects of the coefficients obtained with our method during the trials. the $\star$  symbol corresponds to statistically significant (t-test) nonzero values.}
  \label{fig:ridgebaselineresult}
\end{figure}

### 2019-05-23 paper/synthetic.tex
jr

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds.}
    \label{table:synthetic}
\end{figure}

>>>>

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\textwidth]{synthetic.pdf}
    \caption{Results of synthetic experiments. Each dot depicts the average value of a metric for JRR ($x$-axis) against a competitor ($y$-axis) for each experiment configuration averaged over $20$ random seeds. The prediction error on $Y$ within (in) and outside (out) of the training distribution are weakly but consistantly larger for all of the models as compared to B2B with and without the Sonquist binarization (E). The AUC, measuring the ability of each model to separate causal from noncausal factors, is consistantly larger in B2B as compared to the other models.}
    \label{table:synthetic}
\end{figure}
